### 2014-11-24 Paper_v3.tex
reviewer 3 updates

We calculated control program output and scaled metabolic flux for the positively, uniformly and negatively biased particle swarms (Fig. \ref{fig-flux-pattern}). 
Network A and network C models from the positively (Fig. \ref{fig-flux-pattern}A) and uniformly (Fig. \ref{fig-flux-pattern}B) biased particle swarms showed similar operational patterns, despite differences in kinetic parameters and control structures.
While models from the negatively biased population had error values similar to the correct structures in the previous swarms, they have different flux and control profiles (Fig. \ref{fig-flux-pattern}C). 
In all cases, regardless of network configuration or parameter values, the rate of enzyme decay was small compared to the other fluxes, and all networks had qualitatively similar trends for $E_{3}$ and $E_{6}$ control. 
Moreover, consistent with the correct model structure, production of end product P$_1$ was the preferred branch for all model configurations. 
However, there was variability in P$_{2}$ production flux across the population of models, especially for the uniform swarm when compared with the other cases. 
High P$_{1}$ branch flux resulted in end product inhibition of $E_{1}$ in both network A and network C, however in network D and E, high P$_1$ flux induced $E_1$ activation. 
These trends were manifested in different flux profiles, where the negatively biased population appeared more uniform across the population compared with the other swarms, and had higher $E_{1}$ specific activity. 
Interestingly, the behavior of network A and network C highlighted an artifact of our integration rule; both a positive or negative feedback connection from P$_2$ to $E_{1}$ were ignored because the P$_{1}$ inhibition of $E_{1}$ was dominate. Thus, while theoretically distinct, network A and network C appeared operationally to the PSO algorithm to be that same network. 
On the other hand, networks B, D and E showed distinct behavior that was not consistent with the true network. 
These architectures exhibited either limited inhibition (network B) or activation (network D and E) of $E_{1}$ activity, resulting in significantly different metabolic flux profiles. 
However, the PSO was able to find low error parameter solutions, despite the mismatch in the control structures (error values similar, but not better than the best network A and network C estimates). 
Taken together, these results suggested that a uniform sampling approach could potentially yield an unbiassed estimate of both kinetic parameters and control structures. 
However, the negatively biased particle swarm results illustrated a potential shortcoming of the approach, namely convergence to a local error minimum despite a significantly incorrect control structure. 
This suggested that estimated model structures will need to be further evaluated, for example by generating falsifiable experimental designs which could distinguish between low error solutions.

@@

\subsection*{Formulation and solution of the model equations.}
We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme abundance ($\epsilon_{i}$) in hypothetical cell-free metabolic networks:
\begin{eqnarray}
	\frac{dx_{i}}{dt} & = & \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)\qquad{i=1,2,\hdots,\mathcal{M}}\\
	\frac{d\epsilon_{i}}{dt} & = & -\lambda_{i}\epsilon_{i}\qquad{i = 1,2,\hdots,\mathcal{E}}
\end{eqnarray}where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of metabolites and $\mathcal{E}$ denotes the number of enzymes in the model. 
The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of reaction $j$. 
Typically, reaction $j$ is a non-linear function of metabolite and enzyme abundance, as well as unknown kinetic parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$). 
The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$. 
If $\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$. 
Conversely, if $\sigma_{ij}>0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates metabolite $i$ is not connected with reaction $j$. 
Lastly, $\lambda_{i}$ denotes the scaled enzyme degradation constant. 
The system material balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$ and $\mathbf{\epsilon}\left(t_{o}\right) = \mathbf{1}$ (initially we have 100\% cell-free enzyme abundance).

Each reaction rate was written as the product of two terms, a kinetic term ($\bar{r}_{j}$) and a regulatory term ($v_{j}$):
\begin{equation}\label{eqn:rate-factor}
	r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right) = \bar{r}_{j}v_{j}
\end{equation}
We used multiple saturation kinetics to model the reaction term $\bar{r}_{j}$:
\begin{equation}\label{eqn:rate-bar}
	\bar{r}_{j} = k_{j}^{max}\epsilon_{i}\left(\prod_{s\in{m_{j}^{-}}}\frac{x_{s}}{K_{js} + x_{s}}\right)
\end{equation}where $k_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the scaled enzyme activity which catalyzes reaction $j$, and
$K_{js}$ denotes the saturation constant for species $s$ in reaction $j$. 
The product in Eqn. \eqref{eqn:rate-bar} was carried out over the set of \textit{reactants} for reaction $j$ (denoted as $m_{j}^{-}$).

@@

We minimized the model residual using particle swarm optimization (PSO) \citep{PSO}.
PSO uses a \textit{swarming} metaheuristic to explore parameter spaces. 
A strength of PSO is its ability to find the global minimum, even in the presence of potentially many local minima, by communicating the local
error landscape experienced by each particle collectively to the swarm. Thus, PSO acts both as a local and a global search algorithm. 
For each iteration, particles in the swarm compute their local error by evaluating the model equations using their specific parameter vector realization.
From each of these local points, a globally best error is identified. Both the local and global error 
are then used to update the parameter estimates of each particle using the rules:
\begin{eqnarray}
	\mathbf{\Delta}_{i} &=&\theta_{1}\mathbf{\Delta}_{i} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{k}_{i}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G} - \mathbf{k}_{i}\right) \\
	\mathbf{k}_{i} &=& \mathbf{k}_{i} + \mathbf{\Delta}_{i}
\end{eqnarray}where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes local best solution found by particle $i$, and
$\mathcal{G}$ denotes the best solution found over the entire population of particles. The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
parameters ($\mathcal{K}\times{1}$). In thus study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$. The quality of parameter
estimates was measured using two criteria, goodness of fit (model residual) and angle between the estimated parameter vector $\mathbf{k}_{j}$ and the true parameter set $\mathbf{k}^{*}$:
\begin{equation}
	\alpha_{j} = cos^{-1}\left(\frac{\mathbf{k}_{j}\cdot{\mathbf{k^{*}}}}{\norm{\mathbf{k}_{j}}\norm{\mathbf{k^{*}}}}\right)
\end{equation}If the candidate parameter set $\mathbf{k}_{j}$ were perfect, the residual between the model and synthetic data 
and the angle between $\mathbf{k}_{j}$ and the true parameter set $\mathbf{k}^{*}$ would be equal to zero.

@@

\bibliography{References_v2}

@@

\begin{figure}
\centering
\includegraphics[width=0.50\textwidth]{./figs/Figure-4-OnOffSimulations.pdf}
\caption{ON/OFF control simulations for network A and network B for an ensemble of kinetic parameter sets versus time (N = 100). 
For each case, N = 100 simulations were conducted using kinetic and initial conditions generated randomly from a hypothetical true parameter set. 
The gray area represents $\pm$ one standard deviation surrounding the mean. 
Control parameters were fixed during the ensemble calculations.
\textbf{A}: End product P$_{1}$ abundance versus time for Network A. 
The abundance of P$_{1}$ decreased with end product inhibition of $E_{1}$ activity (Control-ON) versus the no inhibition case (Control-OFF). 
\textbf{B}: End product P$_{2}$ abundance versus time for Network B. Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{2}$ abundance (Control-ON) versus the
no inhibition case (Control-OFF).
\textbf{C}: End product P$_{1}$ abundance versus time for Network A. 
Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{1}$ abundance (Control-ON) versus the no inhibition case (Control-OFF).
}\label{fig-onoff-simulations}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-5-ParameterFit.pdf}
\caption{Parameter estimation from synthetic data for the same and mismatched allosteric control logic using particle swarm optimization (PSO). 
Synthetic experimental data was generated from a hypothetical parameter set using Network A, 
where substrate $S$, end product P$_{1}$ and intermediate M$_5$ were sampled approximately every 20 minutes. 
For cases $\textbf{A,B}$ 20 particles were initialized with randomized parameters and allowed to search for 300 iterations. 
\textbf{A,B}: PSO estimated an ensemble of parameters sets (N = 20) consistent with the synthetic experimental data assuming the correct
enzymatic and control connectivity starting from randomized initial parameters.
\textbf{C,D}: In the presence of control mismatch (Network B control policy simulated with Network A kinetic parameters) 
the ensemble of models did not describe the synthetic data. 
}\label{fig-parameter-fit}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{./figs/Figure-7-ControlSearch.pdf}
\caption{Combined control and kinetic parameter search using modified particle swarm optimization (PSO). 
A population of N = 100 particles was initialized with randomized kinetic parameters and one of five possible control configurations (Network A - E).
Simulation error was minimized for a synthetic data set 
($S$, end product P$_{1}$ and intermediate M$_5$ sampled approximately every 20 min) generated using Network A.
\textbf{A}: Simulation error versus parameter set angle for N = 100 particles biased toward the correct regulatory program (A,B,C,D,E) = (40\%, 10\%, 20\%, 20\% and 10\%).
\textbf{B}: Simulation error versus parameter set angle for N = 100 uniformly distributed particles (A,B,C,D,E) = (20\%, 20\%, 20\%, 20\% and 20\%).
\textbf{C}: Simulation error versus parameter set angle for N = 100 negatively biased particles (A,B,C,D,E) = (10\%, 40\%, 10\%, 20\% and 20\%).
Network A (the correct structure) was preferentially identified for positively and uniform biased particle distributions, but misidentified in the presence of
a large incorrect bias. 
}\label{fig-control-search}
\end{figure}

>>>>

We calculated control program output and scaled metabolic flux for the positively, uniformly and negatively biased particle swarms (Fig. \ref{fig-flux-pattern}). 
Network A and network C models from the positively (Fig. \ref{fig-flux-pattern}A) and uniformly (Fig. \ref{fig-flux-pattern}B) biased particle swarms showed similar operational patterns, despite differences in kinetic parameters and control structures.
While models from the negatively biased population had error values similar to the correct structures in the previous swarms, they have different flux and control profiles (Fig. \ref{fig-flux-pattern}C). 
In all cases, regardless of network configuration or parameter values, the rate of enzyme decay was small compared to the other fluxes, and all networks had qualitatively similar trends for $E_{3}$ and $E_{6}$ control. 
Moreover, consistent with the correct model structure, production of end product P$_1$ was the preferred branch for all model configurations. 
However, there was variability in P$_{2}$ production flux across the population of models, especially for the uniform swarm when compared with the other cases. 
High P$_{1}$ branch flux resulted in end product inhibition of $E_{1}$ in both network A and network C, however in network D and E, high P$_1$ flux induced $E_1$ activation. 
These trends were manifested in different flux profiles, where the negatively biased population appeared more uniform across the population compared with the other swarms, and had higher $E_{1}$ specific activity. 
Interestingly, the behavior of network A and network C highlighted an artifact of our integration rule; both a positive or negative feedback connection from P$_2$ to $E_{1}$ were ignored because the P$_{1}$ inhibition of $E_{1}$ dominated. Thus, while theoretically distinct, network A and network C appeared operationally to the PSO algorithm to be the same network. 
On the other hand, networks B, D and E showed distinct behavior that was not consistent with the true network. 
These architectures exhibited either limited inhibition (network B) or activation (network D and E) of $E_{1}$ activity, resulting in significantly different metabolic flux profiles. 
However, the PSO was able to find low error parameter solutions, despite the mismatch in the control structures (error values similar, but not better than the best network A and network C estimates). 
Taken together, these results suggested that a uniform sampling approach could potentially yield an unbiassed estimate of both kinetic parameters and control structures. 
However, the negatively biased particle swarm results illustrated a potential shortcoming of the approach, namely convergence to a local error minimum despite a significantly incorrect control structure. 
This suggested that estimated model structures will need to be further evaluated, for example by generating falsifiable experimental designs which could distinguish between low error solutions.

@@

\subsection*{Formulation and solution of the model equations.}
We used ordinary differential equations (ODEs) to model the time evolution of metabolite ($x_{i}$) and scaled enzyme abundance ($\epsilon_{i}$) in hypothetical cell-free metabolic networks:
\begin{eqnarray}
	\frac{dx_{i}}{dt} & = & \sum_{j = 1}^{\mathcal{R}}\sigma_{ij}r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)\qquad{i=1,2,\hdots,\mathcal{M}}\\
	\frac{d\epsilon_{i}}{dt} & = & -\lambda_{i}\epsilon_{i}\qquad{i = 1,2,\hdots,\mathcal{E}}
\end{eqnarray}where $\mathcal{R}$ denotes the number of reactions, $\mathcal{M}$ denotes the number of metabolites and $\mathcal{E}$ denotes the number of enzymes in the model. 
The quantity $r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right)$ denotes the rate of reaction $j$. 
Typically, reaction $j$ is a non-linear function of metabolite and enzyme abundance, as well as unknown kinetic parameters $\mathbf{k}$ ($\mathcal{K}\times{1}$). 
The quantity $\sigma_{ij}$ denotes the stoichiometric coefficient for species $i$ in reaction $j$. 
If $\sigma_{ij}>0$, metabolite $i$ is produced by reaction $j$. 
Conversely, if $\sigma_{ij}<0$, metabolite $i$ is consumed by reaction $j$, while $\sigma_{ij} = 0$ indicates metabolite $i$ is not connected with reaction $j$. 
Lastly, $\lambda_{i}$ denotes the scaled enzyme degradation constant. 
The system material balances were subject to the initial conditions $\mathbf{x}\left(t_{o}\right) = \mathbf{x}_{o}$ and $\mathbf{\epsilon}\left(t_{o}\right) = \mathbf{1}$ (initially we have 100\% cell-free enzyme abundance).

Each reaction rate was written as the product of two terms, a kinetic term ($\bar{r}_{j}$) and a regulatory term ($v_{j}$):
\begin{equation}\label{eqn:rate-factor}
	r_{j}\left(\mathbf{x},\mathbf{\epsilon},\mathbf{k}\right) = \bar{r}_{j}v_{j}
\end{equation}
We used multiple saturation kinetics to model the reaction term $\bar{r}_{j}$:
\begin{equation}\label{eqn:rate-bar}
	\bar{r}_{j} = k_{j}^{max}\epsilon_{i}\left(\prod_{s\in{m_{j}^{-}}}\frac{x_{s}}{K_{js} + x_{s}}\right)
\end{equation}
where $k_{j}^{max}$ denotes the maximum rate for reaction $j$, $\epsilon_{i}$ denotes the scaled enzyme activity which catalyzes reaction $j$, and
$K_{js}$ denotes the saturation constant for species $s$ in reaction $j$. 
The product in Eqn. \eqref{eqn:rate-bar} was carried out over the set of \textit{reactants} for reaction $j$ (denoted as $m_{j}^{-}$).

@@

We minimized the model residual using particle swarm optimization (PSO) \citep{PSO}.
PSO uses a \textit{swarming} metaheuristic to explore parameter spaces. 
A strength of PSO is its ability to find the global minimum, even in the presence of potentially many local minima, by communicating the local
error landscape experienced by each particle collectively to the swarm. Thus, PSO acts both as a local and a global search algorithm. 
For each iteration, particles in the swarm compute their local error by evaluating the model equations using their specific parameter vector realization.
From each of these local points, a globally best error is identified. Both the local and global error 
are then used to update the parameter estimates of each particle using the rules:
\begin{eqnarray}
	\mathbf{\Delta}_{i} &=&\theta_{1}\mathbf{\Delta}_{i} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{k}_{i}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G} - \mathbf{k}_{i}\right) \\
	\mathbf{k}_{i} &=& \mathbf{k}_{i} + \mathbf{\Delta}_{i}
\end{eqnarray}
where $\mathbf{\Delta_i}$ denotes the perturbation to the vector of parameters $\mathbf{k_i}$ for particle $i$. 
$\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes the best local solution found by particle $i$, and
$\mathcal{G}$ denotes the best solution found over the entire population of particles. The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
parameters ($\mathcal{K}\times{1}$). In this study, we used $\left(\theta_{1},\theta_{2},\theta_{3}\right) = \left(1.0, 0.05564, 0.02886\right)$. The quality of parameter
estimates was measured using two criteria, goodness of fit (model residual) and angle between the estimated parameter vector $\mathbf{k}_{j}$ and the true parameter set $\mathbf{k}^{*}$:
\begin{equation}
	\alpha_{j} = cos^{-1}\left(\frac{\mathbf{k}_{j}\cdot{\mathbf{k^{*}}}}{\norm{\mathbf{k}_{j}}\norm{\mathbf{k^{*}}}}\right)
\end{equation}If the candidate parameter set $\mathbf{k}_{j}$ were perfect, the residual between the model and synthetic data 
and the angle between $\mathbf{k}_{j}$ and the true parameter set $\mathbf{k}^{*}$ would be equal to zero.

@@

\bibliography{References_v3}

@@

\begin{figure}
\centering
\includegraphics[width=0.50\textwidth]{./figs/Figure-4-OnOffSimulations.pdf}
\caption{ON/OFF control simulations for Network A and Network B for an ensemble of 100 kinetic parameter sets versus time. 
For each case, simulations were conducted using kinetic and initial conditions generated randomly from a hypothetical true parameter set. 
The gray area represents $\pm$ one standard deviation surrounding the mean. 
Control parameters were fixed during the ensemble calculations.
\textbf{A}: End product P$_{1}$ abundance versus time for Network A. 
The abundance of P$_{1}$ decreased with end product inhibition of $E_{1}$ activity (Control-ON) versus the no inhibition case (Control-OFF). 
\textbf{B}: End product P$_{2}$ abundance versus time for Network B. Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{2}$ abundance (Control-ON) versus the
no inhibition case (Control-OFF).
\textbf{C}: End product P$_{1}$ abundance versus time for Network A. 
Inhibition of branch point $E_{6}$ by end product P$_{1}$ decreased P$_{1}$ abundance (Control-ON) versus the no inhibition case (Control-OFF).
}\label{fig-onoff-simulations}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure-5-ParameterFit.pdf}
\caption{Parameter estimation from synthetic data for the same and mismatched allosteric control logic using particle swarm optimization (PSO). 
Synthetic experimental data was generated from a hypothetical parameter set using Network A, 
where substrate $S$, end product P$_{1}$ and intermediate M$_5$ were sampled approximately every 20 minutes. 
For cases $\textbf{A,B}$ 20 particles were initialized with randomized parameters and allowed to search for 300 iterations. 
\textbf{A,B}: PSO estimated an ensemble of 20 parameters sets consistent with the synthetic experimental data assuming the correct
enzymatic and control connectivity starting from randomized initial parameters.
\textbf{C,D}: In the presence of control mismatch (Network B control policy simulated with Network A kinetic parameters) 
the ensemble of models did not describe the synthetic data. 
}\label{fig-parameter-fit}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{./figs/Figure-7-ControlSearch.pdf}
\caption{Combined control and kinetic parameter search using modified particle swarm optimization (PSO). 
A population of 100 particles was initialized with randomized kinetic parameters and one of five possible control configurations (Network A - E).
Simulation error was minimized for a synthetic data set 
($S$, end product P$_{1}$ and intermediate M$_5$ sampled approximately every 20 min) generated using Network A.
\textbf{A}: Simulation error versus parameter set angle for 100 particles biased toward the correct regulatory program (A,B,C,D,E) = (40\%, 10\%, 20\%, 20\% and 10\%).
\textbf{B}: Simulation error versus parameter set angle for 100 uniformly distributed particles (A,B,C,D,E) = (20\%, 20\%, 20\%, 20\% and 20\%).
\textbf{C}: Simulation error versus parameter set angle for 100 negatively biased particles (A,B,C,D,E) = (10\%, 40\%, 10\%, 20\% and 20\%).
Network A (the correct structure) was preferentially identified for positively and uniform biased particle distributions, but misidentified in the presence of
a large incorrect bias. 
}\label{fig-control-search}
\end{figure}
